# Modeling {-}

In order to predict if income is above or below 50,000 we build predictive models. 

The initial models built include a logistic regression, a random forest, and a naive bayes classifier. Using these models on a training set and testing the results on a testing set, we find which models are initially competititve. 

```{r models, warning = FALSE, message = FALSE}
#splitting the data in to train, validate, and testing groups
s = 1:nrow(flat)
train = sample(s, 12000)
validate = sample(s[!s %in% train], 4000)
test = s[(!s %in% train) & (!s %in% validate)]
n = 10

#creating and testing logisitic regression
l = glm(over_50k ~ ., data = flat[train, ], family = binomial(link = "logit"))
p = predict(l, flat[validate, ], type = "response")
perc = (p>0.05) == (flat[validate, ]$over_50k == "1")
#error of 0.361, not a great predictor.

#creating and testing a random forest as predictor
x = rep(NA, n)
for (i in 1:n) {
  rf = randomForest(over_50k ~ ., mtry = 12, ntree = 30, data = flat, subset = train)
  rf2 = predict(rf, flat[validate,])
  x[i] = 1 - sum(rf2 == flat[validate, ]$over_50k)/length(rf2)
}
#mean error of 0.176, much better

#creating and testing a naive bayes model to predict over_50k
nb = naiveBayes(x= flat[, c(-14)], y = flat$over_50k, subset = train)
p = predict(nb, flat[validate, c(-14)])
t = table(pred=p,true=flat[validate,]$over_50k)
#error of 0.176, same as the random forest.

```

The logisitic regression model has an error rate of `r 1 -  sum(perc)/length(validate)`. The Random Forest has a mean error rate of `r mean(x)`. The Naive Bayes has an error rate of `r 1 - ((t[1, 1] + t[2, 2])/sum(t))`. Given the much lower error rates of the Random Forest and Naive Bayes classifier, tuning the parameters of the forest is likely to give an improved error rate. 

```{r final_model, echo = FALSE}
rf = randomForest(over_50k ~ ., mtry = 3, ntree = 500, data = flat, subset = train)
rf2 = predict(rf, flat[validate,])
```
After some tuning of parameters, the error rate of the model is `r 1 - sum(rf2 == flat[validate, ]$over_50k)/length(rf2)`